---
title: "Ch1 - Data Viz with ggplot2"
output:
  html_document:
    toc: true
    toc_float: true
---

# Introduction

Load `tidyverse` library. Install if needed.


```{r}
#install.packages("tidyverse")
library(tidyverse)
```

Remember, if needed, can always call a function within a specific package
by using `package::function()`.

# First Steps

Do cars with large engines use more fuel than cars with small engines?

* what does this relationship look like?
* is it linear, nonlinear?

## The `mpg` Data Frame

We'll use the `mpg` dataset that comes with `ggplot2`.

```{r}
mpg
```

What kind of data structure is `mpg`?

```{r}
class(mpg)
```

For now, let's not worry about tibbles vs data frames. Let's just
call this thing a data frame. It's a table of data with each observation
in a row and each attribute in a column. Attributes have data types.

```{r}
str(mpg)
```

Learn more about this dataset.

```{r}
?mpg
```

## Creating a ggplot

Let's scatter displacement (X axis) vs hwy (Y axis).

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Clearly, highway mpg decreases as displacement increases. The decreases appears
somewhat non-linear.

Notice the structure of the `ggplot` command.

* specify a dataset with `ggplot(data = mpg)`
* add a layer containing the points through `geom_point(mapping = aes(x=displ, y=hwy))`
* the `geom_point` layer provides a mapping between the variables and visual properties of the plot.
* the `aes` stands for aesthetic and maps specific axes to variables

## A graphing template

We can make a general template that we can reuse.

```{r eval=FALSE}
ggplot(data = <DATA>) + 
  geom_point(mapping = aes(<MAPPINGS>))
             
```
## Exercises

** Ex 1 **

```{r}
ggplot(data = mpg)
```

We get an empty graph "container".

** Ex 2 **

How many rows in `mtcars`?

```{r}
nrow(mtcars)
```

How many columns?

```{r}
ncol(mtcars)
```

We can confirm this using the `str` command.
```{r}
str(mtcars)
```

** Ex 3 **

```{r}
summary(mpg$drv)
table(mpg$drv)
```

According to `?mpg`, `drv` uses "f = front-wheel drive, r = rear wheel drive, 4 = 4wd" as factor values.

** Ex 4 **

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy))
             
```


Since `cyl` only takes on a few unique values, we get this "banding effect". Not
surprisingly, there does seem to be a decrease in `hwy` as `cyl` increases since
more cyclinders is likely associated with greater values of displacement. We can
check that.


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = displ))

```


# Aesthetic Mappings

So far we mapped variables to positions in space via an x-y coordinate system.
We can also map variables to other aesthetics such as color, shape and size. For
example, `class` gives the "type" of car. Let's map it to color. When we do
this, `ggplot2` automatically asigns colors to each unique value of the `class`
variable. What do you think would happen if you map color to a continuous
variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

Notice how the 2-seaters cluster. Your interpretation?

Let's try transmission as the color.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = trans))
```
Hmm, not very enlightening.

Now map class to size. You'll get a warning. What do you think it's telling you and why?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Now map `class` to alpha (transparency).

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

Try shape. Read the warning - it's important.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Consider the following example. What are two important differences in how
`colour =` is used?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), colour = "blue")
```

## Exercises

** Ex 1 **

Why aren't the points blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = "blue"))
```

** Ex 2 **

Which variables in `mpg` are categorical and which are continuous?

```{r}
str(mpg)
```

```{r}
summary(mpg)
```

** Ex 3 **

Map a continuous variable to color, size and shape.

We've already done this for color and saw that we got a color gradient.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cty))
```

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = cty))
```
** Ex 4 **

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = hwy))
```


** Ex 5 **

What does the `stroke` aesthetic do?

```{r}
?geom_point
```

Ah, it's for the width of border on a shape.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)

```

Wow, that's one ugly plot. Notice where the `aes()` function is placed? Implications?

** Ex 6 **

Map color to something other than a variable.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ > 5))
```

## Facets

To create "small multiples" we use *faceting*. For one faceting variable,
we use `facet_wrap`. The argument is an R formula. When you see
`~ class`, say "by class".

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class)
```


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow=2)
```

To facet by two variables, use `facet_grid`.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)
```

## Exercises

** Ex 1 **

What if you facet on a continuous variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = hwy)) +
  facet_wrap(~ displ)
```

** Ex 2 **

No point in plot below means no such combination. Same idea as to why you
might have empty plots in a facet plot.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```

** Ex 3 **

What does the dot do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```

The dot lets you create a row or column of plots faceted by a single variable.

** Ex 4 **

Consider the first faceted plot again. Compare to using `color = class`. 
Advantages and disadvantages of each?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow=2)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) 
```

## Geometric objects

A *geom* is the geometric object (line, bar, point, ...) used to represent
the data in a ggplot.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Some additional aesthetic mappings are possible for certain geoms.

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

We can make the plot above even easier to follow by mapping `drv` to color
and including the raw data. Notice
that we are putting multiple geoms on the same plot.

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) +
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))
```

Notice however that this makes it cumbersome to change, say the x variable, since
you'd need to change it in two places - within the smooth and point geoms. What
we want is the aesthetic mapping to be "global". You can probably guess how
you'd do that.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth() +
  geom_point()
```

Now we can add specific mappings to the geoms.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(linetype = drv)) +
  geom_point(aes(color = drv))
```

Try to pull that one off in Excel. :)

You can even use different data in each geom. Notice how the standard error
of the smoothing line is suppressed by the `se = FALSE` argument to `geom_smooth()`.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) +
  geom_point(aes(color = class))
```

The **ggplot2** cheat sheet is helpful.

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

For geoms that use a single geometric object to represent multiple variable
values (such as a line), you can set a `group` aesthetic to get multiple
instances of the geom. You should use a discrete variable for the `group`.

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv), show.legend = TRUE)
```

No legend. Instead, we need to map a discrete variable to another aesthetic such
as color.

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv, group = drv))
```

## Exercises

** Ex 1 **

Line plot doesn't really make sense with these variables but that's ok.

```{r}
ggplot(data = mpg) + 
  geom_line(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = drv, y = hwy))
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth() +
  geom_point()
```

```{r}
ggplot() + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy))
```

## Statistical transformations

Consider a bar chart.

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut))
```

So, by default, the height of each bar corresponds to the number
of records in the categorical variable values on the x-axis. The
default statistic is count. If you have a variable which you want to
use as the height of the bar, you change the stat to what is known
as "identity" - the value itself.

```{r}
demo <- tribble(
  ~a,    ~b,
  "bar_1", 20,
  "bar_2", 30,
  "bar_3", 40
)
```

```{r}
demo
```

```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = a, y = b), stat = "identity")
```

If we use a categorical variable with `stat_count`, we get a bar
chart by default.

```{r}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

A histogram also produces a bar chart after binning the variable of interest.

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = hwy))
```

The y-axis are counts. If you want proportions instead, we can set
the y aesthetic to one of the "computed variables" created by `ggplot`.


```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = hwy, y = ..density..))
```

Another way is using `geom_bar` and `y = ..prop..`.

```{r}
ggplot(data = mpg) + 
  geom_bar(aes(x = hwy, y = ..prop..))
```

You can use `stat_summary` to build a more complex plot which displays
multiple computed values on the y-axis. For example, here's a plot
showing median values along with bars to show the range.

```{r}
ggplot(data = diamonds) +
  stat_summary(mapping = aes(x = cut, y = depth),
               fun.ymax = max,
               fun.ymin = min,
               fun.y = median)
```

### Exercises

** Ex 1 **

The default geom for `stat_summary` is called a *pointrange*. See `?stat_summary`.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = depth)) +
  stat_summary()
```

To reproduce the `stat_summary`, we need to do the upfront work of computing
the `ymin` and `ymax` values. We'll use `dplyr` for this.

```{r}
depth_summary <- diamonds %>% 
  group_by(cut) %>%
  summarize(med_depth = median(depth),
            max_depth = max(depth),
            min_depth = min(depth)
  )

ggplot(data = depth_summary) +
  geom_pointrange(mapping = aes(x = cut, y = med_depth,
                      ymax = max_depth, ymin = min_depth))
```

Could also just set `stat = "summary"` in `geom_pointrange()`.

```{r}
ggplot(data = diamonds) +
  geom_pointrange(mapping = aes(x = cut, y = depth),
                  stat = "summary",
                  fun.ymin = min,
                  fun.ymax = max,
                  fun.y = median)
```

** Ex 2 **

`geom_col` vs `geom_bar`

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut))
```

Whereas `geom_bar` uses count of cases in bin as bar height, `geom_col` uses
a given y value for bar height. I'll use mean of `depth`.

```{r}
ggplot(data = diamonds) +
  geom_col(aes(x = cut, y=mean(depth)))
```

## Positioning

What happens if we set color aesthetic to categorical variable?

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, color = cut))
```

Hmm, not quite satisfying. Set `fill` property instead.

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, fill = cut))
```

Now, if we use a different categorical variable for `fill`, we get a 
stacked bar chart. Nice.

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, fill = clarity))
```

It's actually the *position* argument that causes the stacking. The default value for
`geom_bar` is `position = "stack"`. Other options include "identity", "fill", and "dodge". Let's see what "fill"" does as it's the most relevant for bars.

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, fill = clarity), position = "fill")
```

The "dodge" value gives clustered bar chart.

```{r}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, fill = clarity), position = "dodge")
```


Another position value is called "jitter" and is useful with scatter plots
in which there are lot of duplicate values (which then end up overlapping into
a single point). 

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy), position = "jitter")
```

This is so useful that **ggplot2** comes with a `geom_jitter` as a shortcut.

```{r}
ggplot(data = mpg) +
  geom_jitter(aes(x = displ, y = hwy))
```

### Exercises

** Ex 1 **

What's wrong with this? Improve.

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = cty, y = hwy))
```
Well, it's banded because of the precision of measurement. I'm not so sure
this is "bad". If we jitter the plot, it gives impression that values were
measured at a greater level of precision than they actually were. On the other hand, we also lose a sense of how many (x,y) pairs
there are at each plotted point since they overlap. See Exercise 3.

```{r}
ggplot(data = mpg) +
  geom_jitter(aes(x = cty, y = hwy))
```

** Ex 2 **

How to exercise fine control over jittering. See `?geom_jitter.

Looks like `width` and `height` properties are available to fine tune jitter.

** Ex 3 **

`geom_jitter` vs `geom_count`

By using `geom_count` we keep the banding but get a sense of how
many pairs there are at each plotted by point using point size. It
was designed to deal with "overplotting".

```{r}
ggplot(data = mpg) +
  geom_count(aes(x = cty, y = hwy))
```

** Ex 4 **

Default position adjustment for boxplot? By default, `position = "dodge"`. 

```{r}
ggplot(data = mpg) +
  geom_boxplot(aes(x = drv, y = cty))
```

What happens if we use "identity"? I imagine we'll get overlapping boxplots.

```{r}
ggplot(data = mpg) +
  geom_boxplot(aes(x = drv, y = cty), position = "identity")
```

Nope.

## Coordinate systems

We are usually just using the default Cartesion X-Y coordinate system. You can rotate it so that X is vertical using `coord_flip`. It's helpful when you have
long labels too.

```{r}
ggplot(data = mpg) +
  geom_boxplot(aes(x = class, y = cty)) +
  coord_flip()
```


Other more complicated coordinate systems such as polar coordinates and mapping related coordinate systems are also available. Let's do a quick map and also
see how you can save intermediate plot objects and then add on to them.

```{r}
nz <- map_data("nz")
str(nz)
```

```{r}
nz_map <- ggplot(data=nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "blue", color = "black")

nz_map
```

```{r}
nz_map +
  coord_quickmap()
```

A short polar coordinates example shows link between a bar chart and something known
as a Coxcomb plot.

```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(
    mapping = aes(x = cut, fill = cut),
    show.legend = FALSE,
    width = 1
  ) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

```

```{r}
bar
```

```{r}
bar + coord_flip()
```

```{r}
bar + coord_polar()
```

### Exercises

** Ex 1 **

Turn a stacked bar chart into a pie chart. 

First we need a single stacked bar (i.e. one x category value).

```{r}
gg_stckbar <- ggplot(data = diamonds) +
  geom_bar(aes(x = factor(1), fill = clarity))

gg_stckbar
```

```{r}
gg_stckbar + coord_polar()
```

Not a pie chart. From `?coord_polar` we see that the argument `theta` specified whether
the x or y variable should map to the angle in the coordinate system. For a pie we need
to set `theta = "y"` (default is "x")

```{r}
gg_stckbar + coord_polar(theta = "y")
```


We haven't touched on it yet, but there's so much more that can be done
with **ggplot2** in terms of plot styles.

```{r}
gg_stckbar + 
  coord_polar(theta = "y") + 
  scale_fill_brewer(palette="Blues") +
  theme_minimal() +
```


** Ex 2 **

What does `labs()` do? What about a graph title?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) +
  labs(x = "Engine displacement", y = "MPG (highway)") +
  ggtitle("Engine displacement vs gas mileage") 
  
```

Can also set title, subtitle and caption via `labs()`.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) +
  labs(x = "Engine displacement", 
       y = "MPG (highway)",
       title = "Engine displacement vs gas mileage",
       subtitle = "Does car class matter?",
       caption = "Source: See ?mpg"
  )

  
```

```{r}
library(ggthemes)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) +
  labs(x = "Engine displacement", 
       y = "MPG (highway)",
       title = "Engine displacement vs gas mileage",
       subtitle = "Does car class matter?",
       caption = "Source: See ?mpg"
  ) +
  theme_tufte()

  
```

** Ex 3 **

`coord_quickmap()` val `coord_map()`?

From reading the help, `coord_quickmap()` relies on a "quick approximation" for 
computing the map projection. the approximation works better near the equator
according to the docs.

```{r}
nz_map +
  coord_quickmap()
```

```{r}
nz_map +
  coord_map()
```

** Ex 4 **

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_point()
```

An *abline* gives a 45 degree reference line or Y=X.

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline()
```

To ensure that units on both axes are the same size, we can use `coord_fixed()`.

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed()
```